Processing large surveys with package MRP-KIT

### 2021-07-03 case-study/Munging_CCES.Rmd
completed mapping discussion

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE, prompt=TRUE)
options(digits=2)
if (!require("ipumsr")) stop("This case study requires ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
if (!require("labelled")) stop("This case study requires labelled package. It can be installed using the following command: install.packages('labelled')")
library(ipumsr)
library(labelled)
```

@@

The Common Content portion of the survey consists of questions asked of all subjects.
Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, race, and geographic location, among others.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```

@@

```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")
demo_cols <- c(which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "inputstate"))
# don't allow missing demographics
keep_rows <- complete.cases(cces_2018[, demo_cols])

@@

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
ipums_val_labels(acs_2018_subset$AGE)
```

@@

@@

#### 1-1 mapping on a long list: state of residence, 51 categories

@@

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories as follows.

@@

Mpping "Some College" to both 1 year and 3 years of college is also possible,
reasoning that the goal of a 4-year college degree wasn't obtained.
The choice depends on both an understanding of the survey data and the particular goals
of your analysis.
MRP-KIT can only provide the mechanism for recording and carrying out the mapping.
```{r}
educ_mapping = list(
    "No HS" = "N/A or no schooling",
    "No HS" = "Nursery school to grade 4",
    "No HS" = "Grade 5, 6, 7, or 8",
    "No HS" = "Grade 9",
    "No HS" = "Grade 10",
    "No HS" = "Grade 11",
    "High School Graduate" = "Grade 12",
    "Some College" = "1 year of college",
    "2-year" = "2 years of college",
    "2-year" = "3 years of college",
    "4-year" = "4 years of college",
    "Post-grad" = "5+ years of college"
)
q_educ <- SurveyQuestion$new(name = "educ",
    col_names = c("educ", "EDUC"),
    values_map= educ_mapping)
q_educ$name()
q_educ$col_names()
q_educ$values()
```

@@

We propose to add a column to the ACS survey which combines information present in the `RACE` and `HISPAN` variables.
```{r}
# create column which blends "race" "hispan"
acs_race2 <- as_factor(acs_2018_subset$RACE)

# population percentages by label
prop.table(table(acs_race2))

# if "hispan" is "Yes", "race" label is "Hispanic".
is_hispan <- acs_2018_subset$HISPAN > 0 & acs_2018_subset$HISPAN < 5
length(is_hispan[is_hispan == TRUE]) / length(is_hispan)
levels(acs_race2) = c(levels(acs_race2), "Hispanic")
acs_race2[is_hispan==TRUE] <- "Hispanic"

# adjusted percentages
prop.table(table(acs_race2))

# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$race2 = to_labelled(acs_race2)
```

A further possibility is to similarly combine the CCES variables "race" and "hispan" 
so that the question of "Hispanic origin" is "Yes", the race-hispan label is "Hispanic".
Were we to do this, we would create a new variable "race2" and add it to the
CCES subset so that we could evaluate both classifications.
```{r}
# Population percentages in CCES variable "race"
prop.table(table(cces_2018_subset$race))

# Intersection (and) vs. union (or) of race == "Hispanic" and hispanic == "Yes"
length(which(cces_2018_subset$race == "Hispanic" & cces_2018_subset$hispan == "Yes"))
length(which(cces_2018_subset$race == "Hispanic" | cces_2018_subset$hispan == "Yes"))

# create column which adjusts race labels
cces_race2 <- cces_2018_subset$race
cces_race2[cces_2018_subset$hispan == "Yes"] <- "Hispanic"
prop.table(table(cces_race2))

# add to acs_2018_subset data.frame - respect column types
cces_2018_subset$race2 = cces_race2
```

@@

For the 2018 CCES survey, the marital status of all respondants is known,
therefore categories "skipped" and "not asked" are not problematic.
The ACS variable ["MARST"](https://usa.ipums.org/usa-action/variables/MARST#comparability_section)
has two categories for corresponding to CCES category "Married":
"Married, spouse present", "Married, spouse absent".
Mapping both ACS categories to CCES category "Married" is a one-to-many mapping
which is allowed by MRP-KIT.
The CCES has category "Domestic / civil partnership",
for which there is no corresponding ACS category.
Mapping both partnerships and "Never married" to the ACS category "Never married/single"
is a many-to-one mapping from CCES to ACS, also allowed by MRP-KIT.
However, if we map CES category  "Domestic / civil partnership"
to either ACS category "Married, spouse present" or "Married, spouse absent",
we create a many-to-many mapping.

OK             |                | NOT OK         | 
-------------- | -------------- | -------------- | --------------
CCES | ACS | CCES | ACS
Married | Married, spouse present |Married | Married, spouse present 
Married | Married, spouse absent |Married | Married, spouse absent 
Separated | Separated |Separated | Separated 
Divorced | Divorced |Divorced | Divorced 
Widowed | Widowed |Widowed | Widowed 
Never married | Never married/single |Never married | Never married/single 
*Domestic / civil partnership* | *Never married/single* | *Domestic / civil partnership* | **Married, spouse present**

```{r, error=TRUE}
# This is OK
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Never married"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
q_marstat$values()

# This is NOT
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Married, spouse present"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)

```
Here we propose to add yet another column to the ACS which conflates "Married, spouse present" and "Married, spouse absent".

@@

# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$marst2 = to_labelled(droplevels(acs_marst2))
ipums_val_labels(acs_2018_subset$marst2)
```
Now we can create a mapping which maps both "Married" and ""Domestic / civil partnership"
to the the (new) ACS label "Married".^[
Alternatively, we could try to access more ACS variables which account for
same sex marriages and civil unions and use this information to code new variable "marst2".]

@@

#### Aligned demographics

We have created 5 questions:  `q_sex`, `q_state`, `q_educ`, `q_race`, `q_marstat`.
This will allow us to slice and dice our surveys to a ridiculous level of granularity.
Excelsior!

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE, prompt=TRUE, highlight=FALSE)
options(digits=2)
if (!require("ipumsr")) stop("This case study requires ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
if (!require("labelled")) stop("This case study requires labelled package. It can be installed using the following command: install.packages('labelled')")
library(ipumsr)
library(labelled)
```

@@

The Common Content portion of the survey consists of questions asked of all subjects.

Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, race, and geographic location, among others.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```

@@

```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")

demo_cols <- c(which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "inputstate"))
keep_rows <- complete.cases(cces_2018[, demo_cols]) # no missing demographics

@@

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
```

@@

The CCES is a survey of _national adults_, i.e., eligible
voters over the age of 18 but the ACS is a survey over the entire
US population, including non-citizens.
In order to use the ACS for poststratification, we drop all
records for individuals under the age of 18 and non-citizens.

```{r}
table(as_factor(acs_2018_subset$CITIZEN))
# drop underage, non-citizens
citizen = as_factor(acs_2018_subset$CITIZEN) != "Not a citizen"
over_18 = acs_2018_subset$AGE > 17
acs_2018_subset = acs_2018_subset[(over_18 & citizen) , ]  
```

@@

#### One-to-one mapping over many labels: state of residence, 51 categories

@@

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories so that the CCES label "2-year"
maps to the ACS labels for 2 and 3 years of college.
Mpping "Some College" to both 1 year and 3 years of college is also possible,
on the rationale that the goal was a 4-year college degree, and that it wasn't obtained.
The choice depends on both an understanding of the survey data and the particular goals
of your analysis.

@@

```{r}
educ_mapping = list(
    "No HS" = "N/A or no schooling",
    "No HS" = "Nursery school to grade 4",
    "No HS" = "Grade 5, 6, 7, or 8",
    "No HS" = "Grade 9",
    "No HS" = "Grade 10",
    "No HS" = "Grade 11",
    "High School Graduate" = "Grade 12",
    "Some College" = "1 year of college",
    "2-year" = "2 years of college",
    "2-year" = "3 years of college",
    "4-year" = "4 years of college",
    "Post-grad" = "5+ years of college"
)
q_educ <- SurveyQuestion$new(name = "educ",
    col_names = c("educ", "EDUC"),
    values_map= educ_mapping)
q_educ$name()
q_educ$col_names()
q_educ$values()
```

@@

We propose to add a column to the ACS survey which combines information present in the `RACE` and `HISPAN` variables.
```{r}
# create column which blends "race" "hispan"
acs_race2 <- as_factor(acs_2018_subset$RACE)
# population percentages by label
prop.table(table(acs_race2))
# if "hispan" is "Yes", "race" label is "Hispanic".
is_hispan <- acs_2018_subset$HISPAN > 0 & acs_2018_subset$HISPAN < 5
length(is_hispan[is_hispan == TRUE]) / length(is_hispan)
levels(acs_race2) = c(levels(acs_race2), "Hispanic")
acs_race2[is_hispan==TRUE] <- "Hispanic"
# adjusted percentages
prop.table(table(acs_race2))
# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$race2 = to_labelled(acs_race2)
```

A further possibility is to similarly combine the CCES variables "race" and "hispan" 
so that the question of "Hispanic origin" is "Yes", the race-hispan label is "Hispanic".
Were we to do this, we would create a new variable "race2" and add it to the
CCES subset so that we could evaluate both classifications.
```{r}
# Population percentages in CCES variable "race"
prop.table(table(cces_2018_subset$race))
# Intersection (and) vs. union (or) of race == "Hispanic" and hispanic == "Yes"
length(which(cces_2018_subset$race == "Hispanic" & cces_2018_subset$hispan == "Yes"))
length(which(cces_2018_subset$race == "Hispanic" | cces_2018_subset$hispan == "Yes"))
# create column which adjusts race labels
cces_race2 <- cces_2018_subset$race
cces_race2[cces_2018_subset$hispan == "Yes"] <- "Hispanic"
prop.table(table(cces_race2))
# add to acs_2018_subset data.frame - respect column types
cces_2018_subset$race2 = cces_race2
```

@@

For the 2018 CCES survey, the marital status of all respondants is known,
therefore categories "skipped" and "not asked" are not problematic.
The ACS variable ["MARST"](https://usa.ipums.org/usa-action/variables/MARST#comparability_section)
has two categories for corresponding to CCES category "Married":
"Married, spouse present", "Married, spouse absent".

The CCES has category "Domestic / civil partnership",
for which there is no corresponding ACS category.
We could try to access additional columns of the ACS, as there
is some information on same-sex partnerships and civil unions.
Otherwise, we have a choice of mapping partnerships to either
of the ACS married categories, or to the "Never married/single" category.

If we do the former, we create a many-to-many mapping:
CCES category "Married" maps to two ACS categories: spouse present/absent;
ACS category "Married, spouse present" maps to two CCES categories: married and partnerships.
This is not allowed by MRP-KIT.
If we do the latter, we create two many-to-one mappings:
CCES category "Married" maps to two ACS categories: spouse present/absent;
ACS category "Never married/single" maps to two CCES categories:
never married and civil partnership.  This perfectly OK.

*question: implication of many-to-one mappings between the survey population and the
target population - do we get separate postratification cells?*

OK             |                | NOT OK         | 
-------------- | -------------- | -------------- | --------------
**CCES** | **ACS** | **CCES** | **ACS**
Never married | Never married/single |Never married | Never married/single 
*Domestic / civil partnership* | *Never married/single* | *Domestic / civil partnership* | **Married, spouse present**
Married | Married, spouse present |Married | Married, spouse present 
Married | Married, spouse absent |Married | Married, spouse absent 
Separated | Separated |Separated | Separated 
Divorced | Divorced |Divorced | Divorced 
Widowed | Widowed |Widowed | Widowed

```{r, error=TRUE}
# OK
marstat_mapping = list(
    "Never married" = "Never married/single",
    "Domestic / civil partnership" = "Never married/single",
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
q_marstat$values()

# NOT OK
marstat_mapping = list(
    "Never married" = "Never married/single",
    "Domestic / civil partnership" = "Married, spouse present",
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)

```
Since the distinction in the ACS data between spouse present and spouse absent isn't useful,
we can solve the many-to-many mapping problem by adding a new column to the ACS
which conflates "Married, spouse present" and "Married, spouse absent".

@@

# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$marst2 = to_labelled(droplevels(acs_marst2))
ipums_val_labels(acs_2018_subset$marst2)
```
Now we can create a mapping which maps both "Married" and ""Domestic / civil partnership"
to the the (new) ACS label "Married".^[
Alternatively, we could try to access more ACS variables which account for
same sex marriages and civil unions and combine information across multiple columns, as we did
for ACS variable race2.]

@@

#### Measurement recorded in different units and not binned:  age

The CCES is records age as birthyear but the ACS records age in years.
In both surveys, this is an integer, not a categorical measurement but
MRP-KIT mappings operate over a finite set of categories.

```{r}
acs_birthyear = 2018 - acs_2018_subset$AGE
summary(acs_birthyear)
summary(cces_2018$birthyr)
```
We have to decide how many bins and which years they should span.
We seek to bin the population in ways that respects what we know about
US generational cohorts (citation?), e.g., "baby-boomers" born between 1945 an 1965.

* 7 10-year+ cohorts: 1920-1935, 1936-1945, 1946-1955, 1956-1965, 1966-1975, 1976-1985, 1986-2000
* 4 20-year+ cohorts: 1920-1945, 1946-1965, 1966-1985, 1986-2000

Choosing the former will result in more postratification cells (57% more, to be precise).

### Mapping the survey and target population data with MRP-KIT

We now have mappings for our 6 postratification variables:
`q_sex`, `q_state`, `q_educ`, `q_race`, `q_marstat`, `q_age`;
allowing for (at least) `2 * 51 * 6 * 8 * 6 * 4`  possible subsets of interest.
